Incremental λ-Calculus in Cache-Transfer Style

### 2017-08-22 popl18/source-language.tex
Fix typo.

Our syntax for base terms~$\sterm$ represents $\lambda$-lifted
programs in A'-normal form (\cref{sec:motivating-example}).
We write $\many\tx$ for a sequence of variables of some unspecified length
$\tx_1, \tx_2, \ldots, \tx_m$.
A term can be either a bound variable~$\tx$, or a {\bf let}
binding of $\ty$ in subterm~$\sterm$ to either a new
tuple~$\stuple{\many\tx}$ ($\slet{\ty = \stuple{\many\tx}}\sterm$), or
the result of calling function~$\tf$ on argument~$\tx$
($\slet{\ty = \sapp{\tf}{\tx}}{\sterm}$).  Both~$\tf$ and~$\tx$ are
variables to be looked up in the environment.  Terms cannot contain
$\lambda$-abstractions as they have been $\lambda$-lifted to top-level
definitions, which we encode as closures in the initial environments.
%
Unlike standard ANF we add no special syntax for function
calls in tail position (see \cref{sec:tail-calls} for a discussion
about this limitation).
%
We often inspect the result of
a function call~$\tf\,\tx$, which is not a valid term in our syntax.
To enable this, we write~``$\scall\tf\tx$''
for ``$\slet{\ty = \tf\;\tx}\ty$'' where $\ty$ is
chosen fresh.

@@

We define
the \textit{original environment}~$\oldenv\iectx$
and the \textit{new environment}~$\newenv\iectx$
of a change environment~$\iectx$ by induction over~$\iectx$:
\[
\begin{array}{rcl}
\env\sectxempty & = & \sectxempty \qquad $i = 1, 2$\\
\oldenv{\iectxcons\iectx{\tx = \sclosedvalue; \tdx = \icloseddvalue}}
& = & \iectxcons{\oldenv{\iectx}}{\tx = \sclosedvalue} \\
\newenv{\iectxcons\iectx{\tx = \sclosedvalue; \tdx = \icloseddvalue}}
& = & \iectxcons{\newenv{\iectx}}{\tx = \sclosedvalue \oplus \icloseddvalue}
\end{array}
\]
This last rule makes use of an operation $\oplus$ to update a value
with a change, which may fail at runtime. Indeed, change update
is a partial function written
``$\sclosedvalue \oplus \icloseddvalue$'', defined as follows:
\[
  \begin{array}{rclcl}
    \sclosedvalue & \oplus & \inil & = & \sclosedvalue
    \\
    \sclosedvalue_1 & \oplus & \replace{\sclosedvalue_2} & = & \sclosedvalue_2
    \\
    \sconst & \oplus & \idconst & = & \delta_{\oplus} (\sconst, \idconst)
    \\
    \sclosure{\sectx}{\slam\tx\sterm} & \oplus &
    \iclosure\iectx{\ilam{\tx\,\tdx}{\idterm}} & = &
    \sclosure{(\sectx \oplus \iectx)}{\slam\tx\sterm}
    \\
    \stuple{\sclosedvalue_1, \ldots, \sclosedvalue_n}
    & \oplus
    & \stuple{\icloseddvalue_1, \ldots, \icloseddvalue_n}
    & =
    & \stuple{\sclosedvalue_1 \oplus \icloseddvalue_1, \ldots, \sclosedvalue_n \oplus \icloseddvalue_n}
  \end{array}
\]
\noindent where
\[
    (\tectxcons\sectx{\tx = \sclosedvalue}) \oplus
    (\tectxcons\iectx{\tx = \sclosedvalue; \tdx = \icloseddvalue})
    =
    (\tectxcons{(\sectx \oplus \iectx)}{\tx = (\sclosedvalue \oplus \icloseddvalue)})
\]
Nil and replacement changes can be used to update all values
(constants, tuples, primitives and closures), while tuple changes can
only update tuples, literal changes can only update literals and
closure changes can only update closures. A nil change leaves a value unchanged.
A replacement change overrides the current value~$\sclosedvalue$ with a
new one~$\sclosedvalue'$.
On literals, $\oplus$ is defined via some interpretation
function~$\delta_{\oplus}$.
Change update for a closure
ignores~$\idterm$ instead of combining it with $\sterm$ somehow. This may seem
surprising, but we only need $\oplus$ to behave well for valid changes (as shown
by \cref{lemma:crel-oplus}): for valid closure changes,
$\idterm$ must behave similarly to $\iderive\sterm$ anyway,\pg{Not formal, not
  visible in the definition, but true for a suitable meaning of ``similarly'';
  but I'm clearly not claiming a theorem, just giving a hint for clarity.}
so only environment updates matter. This definition also avoids having to modify
terms at runtime, which would be difficult to implement safely.
\pg{$\oplus$ is not used by derivatives, so commenting this part out
A change application for a closure
ignores~$\idterm$, which may seems strange at first but is needed to
but is needed to
get guarantees about the evaluation of derivatives.}
% With this restriction, we avoid the case of a new
% closure code that would refer to a variable undefined in the closure
% environment.
% \pg{I can imagine fixes to that problem, so I'd propose to stick to the
% stronger arguments.}

>>>>

Our syntax for base terms~$\sterm$ represents $\lambda$-lifted
programs in A'-normal form (\cref{sec:motivating-example}).
We write $\many\tx$ for a sequence of variables of some unspecified length
$\tx_1, \tx_2, \ldots, \tx_m$.
A term can be either a bound variable~$\tx$, or a {\bf let}-binding
of~$\ty$ in subterm~$\sterm$ to either a new
tuple~$\stuple{\many\tx}$ ($\slet{\ty = \stuple{\many\tx}}\sterm$), or
the result of calling function~$\tf$ on argument~$\tx$
($\slet{\ty = \sapp{\tf}{\tx}}{\sterm}$). Both~$\tf$ and~$\tx$ are
variables to be looked up in the environment. Terms cannot contain
$\lambda$-abstractions as they have been $\lambda$-lifted to top-level
definitions, which we encode as closures in the initial environments.
%
Unlike standard ANF we add no special syntax for function
calls in tail position (see \cref{sec:tail-calls} for a discussion
about this limitation).
%
We often inspect the result of
a function call~$\tf\,\tx$, which is not a valid term in our syntax.
To enable this, we write~``$\scall\tf\tx$''
for ``$\slet{\ty = \tf\;\tx}\ty$'' where $\ty$ is
chosen fresh.

@@

We define
the \textit{original environment}~$\oldenv\iectx$
and the \textit{new environment}~$\newenv\iectx$
of a change environment~$\iectx$ by induction over~$\iectx$:
\[
\begin{array}{rcl}
\env\sectxempty & = & \sectxempty \qquad $i = 1, 2$\\
\oldenv{\iectxcons\iectx{\tx = \sclosedvalue; \tdx = \icloseddvalue}}
& = & \iectxcons{\oldenv{\iectx}}{\tx = \sclosedvalue} \\
\newenv{\iectxcons\iectx{\tx = \sclosedvalue; \tdx = \icloseddvalue}}
& = & \iectxcons{\newenv{\iectx}}{\tx = \sclosedvalue \oplus \icloseddvalue}
\end{array}
\]
This last rule makes use of an operation $\oplus$ to update a value
with a change, which may fail at runtime. Indeed, change update
is a partial function written
``$\sclosedvalue \oplus \icloseddvalue$'', defined as follows:
\[
  \begin{array}{rclcl}
    \sclosedvalue & \oplus & \inil & = & \sclosedvalue
    \\
    \sclosedvalue_1 & \oplus & \replace{\sclosedvalue_2} & = & \sclosedvalue_2
    \\
    \sconst & \oplus & \idconst & = & \delta_{\oplus} (\sconst, \idconst)
    \\
    \sclosure{\sectx}{\slam\tx\sterm} & \oplus &
    \iclosure\iectx{\ilam{\tx\,\tdx}{\idterm}} & = &
    \sclosure{(\sectx \oplus \iectx)}{\slam\tx\sterm}
    \\
    \stuple{\sclosedvalue_1, \ldots, \sclosedvalue_n}
    & \oplus
    & \stuple{\icloseddvalue_1, \ldots, \icloseddvalue_n}
    & =
    & \stuple{\sclosedvalue_1 \oplus \icloseddvalue_1, \ldots, \sclosedvalue_n \oplus \icloseddvalue_n}
  \end{array}
\]%
where
\[
    (\tectxcons\sectx{\tx = \sclosedvalue}) \oplus
    (\tectxcons\iectx{\tx = \sclosedvalue; \tdx = \icloseddvalue})
    =
    (\tectxcons{(\sectx \oplus \iectx)}{\tx = (\sclosedvalue \oplus \icloseddvalue)})
\]
Nil and replacement changes can be used to update all values
(constants, tuples, primitives and closures), while tuple changes can
only update tuples, literal changes can only update literals and
closure changes can only update closures. A nil change leaves a value unchanged.
A replacement change overrides the current value~$\sclosedvalue$ with a
new one~$\sclosedvalue'$.
On literals, $\oplus$ is defined via some interpretation
function~$\delta_{\oplus}$.
Change update for a closure
ignores~$\idterm$ instead of combining it with $\sterm$ somehow. This may seem
surprising, but we only need $\oplus$ to behave well for valid changes (as shown
by \cref{lemma:crel-oplus}): for valid closure changes,
$\idterm$ must behave similarly to $\iderive\sterm$ anyway,\pg{Not formal, not
  visible in the definition, but true for a suitable meaning of ``similarly'';
  but I'm clearly not claiming a theorem, just giving a hint for clarity.}
so only environment updates matter. This definition also avoids having to modify
terms at runtime, which would be difficult to implement safely.
\pg{$\oplus$ is not used by derivatives, so commenting this part out
A change application for a closure
ignores~$\idterm$, which may seems strange at first but is needed to
but is needed to
get guarantees about the evaluation of derivatives.}
% With this restriction, we avoid the case of a new
% closure code that would refer to a variable undefined in the closure
% environment.
% \pg{I can imagine fixes to that problem, so I'd propose to stick to the
% stronger arguments.}
